---
title: "Data Exploration"
author: "Christina Chen"
date: "2/22/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)

# data file too large to push to git - split into two halfs in split_data.R
student_data1 = read.csv("els_student_first_half.csv")
student_data2 = read.csv("els_student_second_half.csv")
els = rbind(student_data1, student_data2)
```

# Measures of Success
- standardized test quartile: BYTXCQU 
- standardized test composite: BYTXCSTD
```{r}
# hist(student_data$BYTXCQU)
hist(els$BYTXCSTD)
# some scores below 0 -> should remove
```

- employment income 7 years after graduation (2011): F3ERN2011
```{r}
sum(els$F3ERN2011==0)
hist(log(els$F3ERN2011))
```

- highest level of education earned 8 years after graduation (2012): F3ATTAINMENT
```{r}
ggplot(data=els) +
  geom_bar(aes(x=F3ATTAINMENT))
```

- HS GPA: F1RGPP2

- SAT: TXEESATC

- ACT: TXEEACTC

# Familial
- race: BYRACE
```{r}

```

- family composition: BYFCOMP
- parents' highest level of education: BYPARED
- parents' English fluency
- socioeconomic status: BYSES1, BYSES1QU, BYSES2, BYSES2QU
- family income: BYINCOME
- how far in school parents wants student to go: BYPARASP
- has a computer at home: BYS84C
- has own room: BYS84I
- has Internet access: BYS84D
- parents expect success in school: BYS27I

# Educational (for the school the student attends)
- school type (private/public): BYSCTRL
- school urbanicity: BYURBAN
- dropout prevention program offered: F1A23
- programs for pregnant girls/teenage mothers offered: F1A21F
- individual/family psychological counseling offered: F1A21E
- school-sponsored community service offered: F1A21D
- vocational counseling/services/programs offered: F1A21A
- learning hindered by poor condition of buildings, poor heating/air/light, poor science labs, poor fine arts facilities, poor lack of space, library, lack of texts/supplies, too few computers, lack of multi-media, lack of discipline/safety, poor voc/tech equipment/facilities

BYA50A         14277-14279   Learning hindered by poor condition of buildings
BYA50B         14279-14281   Learning hindered by poor heating/air/light
BYA50C         14281-14283   Learning hindered by poor science labs
BYA50D         14283-14285   Learning hindered by poor fine arts facilities
BYA50E         14285-14287   Learning hindered by lack of space
BYA50F         14287-14289   Learning hindered by poor library
BYA50G         14289-14291   Learning hindered by lack of texts/supplies
BYA50H         14291-14293   Learning hindered by too few computers
BYA50I         14293-14295   Learning hindered by lack of multi-media
BYA50J         14295-14297   Learning hindered by lack of discipline/safety
BYA50K         14297-14299   Learning hindered by poor voc/tech equipment/facilities


- teacher access to technology

BYA41A         14131-14133   Teachers have access to cable TV
BYA41B         14133-14135   Teachers have access to closed-circuit TV
BYA41C         14135-14137   Teachers have access to videodisc player/VCR/DVD
BYA41D         14137-14139   Teachers have access to video camera
BYA41E         14139-14141   Teachers have access to video production studio
BYA41F         14141-14143   Teachers have access to satellite TV hook-up
BYA41G         14143-14145   Teachers have access to videoconferencing equipment
BYA41H         14145-14147   Teachers have access to digital camera
BYA41I         14147-14149   Teachers have access to scanner
BYA41J         14149-14151   Teachers have access to LCD panel
BYA41K         14151-14153   Teachers have access to laptop computer
BYA41L         14153-14155   Teachers have access to Internet
BYA41M         14155-14157   Teachers have access to computer printer

- highest salary paid to full-time teachers: BYA26B
- % full-time teachers that are certified: BYA24A
- % full-time teachers that teach out of field: BYA25A
- percent of 10th graders with free lunch: BY10FLP



```{r}

```

