---
title: "Data Exploration"
author: "Christina Chen"
date: "2/22/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)

# data file too large to push to git - split into two halfs in split_data.R
student_data1 = read.csv("els_student_first_half.csv")
student_data2 = read.csv("els_student_second_half.csv")
els = rbind(student_data1, student_data2)
els_original=els
```

# Measures of Success
- standardized test quartile: BYTXCQU 
- standardized test composite: BYTXCSTD
```{r}
# hist(student_data$BYTXCQU)
hist(els$BYTXCSTD)
# some scores below 0 -> should remove
```

- employment income 7 years after graduation (2011): F3ERN2011
```{r}
sum(els$F3ERN2011==0)
hist(log(els$F3ERN2011))
```

- highest level of education earned 8 years after graduation (2012): F3ATTAINMENT
F3ATTAINMENT
   1 = "No HS credential, no PS attendance"
   2 = "HS credential, no PS attendance"
   3 = "Some PS attendance, no PS credential"
   4 = "Undergraduate certificate"
   5 = "Associates degree"
   6 = "Bachelors degree"
   7 = "Post-Baccalaureate certificate"
   8 = "Master's degree/Post-Master's certificate"
   10 = "Doctoral degree"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
```{r}
els <- els %>% mutate(
  ED_ATNMT = case_when(
    F3ATTAINMENT == 1 ~ "No GED",
    F3ATTAINMENT == 2 ~ "GED",
    F3ATTAINMENT ==3 ~ "PS Attendance",
    F3ATTAINMENT ==4 ~ "Undergrad Cert.",
    F3ATTAINMENT ==5 ~ "Associates",
    F3ATTAINMENT ==6 ~ "Bachelors",
    F3ATTAINMENT ==7 ~ "Post-Bacc. Cert.",
    F3ATTAINMENT ==8 ~ "Masters",
    F3ATTAINMENT ==10 ~ "PhD"
  )
)
els$ED_ATNMT = factor(els$ED_ATNMT, levels=c( "No GED", "GED", "PS Attendance", "Undergrad Cert.", "Associates", "Bachelors", "Post-Bacc. Cert.", "Masters", "PhD"))

els %>%
  filter(!is.na(ED_ATNMT)) %>%
  ggplot() +
    geom_bar(aes(x=ED_ATNMT))
```

```{r}
els %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=F3ERN2011))
```

<!-- - HS GPA: F1RGPP2 -->

<!-- - SAT: TXEESATC -->

<!-- - ACT: TXEEACTC -->

# Familial
- race: BYRACE
BYRACE
   1 = "Amer. Indian/Alaska Native, non-Hispanic"
   2 = "Asian, Hawaii/Pac. Islander,non-Hispanic"
   3 = "Black or African American, non-Hispanic"
   4 = "Hispanic, no race specified"
   5 = "Hispanic, race specified"
   6 = "More than one race, non-Hispanic"
   7 = "White, non-Hispanic"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
```{r}
# AI/AN = American Indian/Alaska Native
# API = Asian, Hawaii/Pacific Islander
els <- els %>% mutate(
  RACE = case_when(
    BYRACE == 1 ~ "AI/AN",
    BYRACE ==2 ~ "API",
    BYRACE ==3 ~ "Black",
    BYRACE ==4 | BYRACE ==5 ~ "Hispanic",
    BYRACE ==6 ~ "Mixed",
    BYRACE ==7 ~ "White"
  )
)
els$RACE = factor(els$RACE, levels=c( "AI/AN", "Mixed", "API", "Black", "Hispanic","White"))

els %>%
  filter(!is.na(RACE)) %>%
  ggplot() +
  geom_bar(aes(x=RACE))

els$RACE = factor(els$RACE, levels=c("Black", "Hispanic", "AI/AN", "Mixed", "API", "White"))
              
els %>%
  filter(!is.na(RACE)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=RACE))

els %>%
  filter(!is.na(RACE)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=RACE))

els %>%
  filter(!is.na(RACE), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=RACE, fill=ED_ATNMT), position = "fill")

sum(is.na(els$RACE))
```

- family composition: BYFCOMP

BYFCOMP
   1 = "Mother and father"
   2 = "Mother and male guardian"
   3 = "Father and female guardian"
   4 = "Two guardians"
   5 = "Mother only"
   6 = "Father only"
   7 = "Female guardian only"
   8 = "Male guardian only"
   9 = "Lives with student less than half time"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
   -9 = "Missing"
```{r}
els <- els %>% mutate(
  BYFCOMP = na_if(BYFCOMP, BYFCOMP<0),
  FAM_COMP = case_when(
    BYFCOMP == 1 | BYFCOMP == 2 | BYFCOMP == 3 | BYFCOMP == 4 ~ "Two Parents",
    BYFCOMP ==5 | BYFCOMP == 6 | BYFCOMP ==7 | BYFCOMP == 8 ~ "Single Parent",
    BYFCOMP == 9 ~ "Absent Parent"
  )
)
# els$RACE = factor(els$RACE, levels=c( "Amer. Indian/Alaska Native", "Mixed", "Asian, Hawaii/Pac. Islander", "Black", "Hispanic","White"))


els %>%
  filter(!is.na(FAM_COMP)) %>%
  ggplot() +
  geom_bar(aes(x=FAM_COMP))

# els$RACE = factor(els$RACE, levels=c("Black", "Hispanic", "Amer. Indian/Alaska Native", "Mixed", "Asian, Hawaii/Pac. Islander", "White"))
              
els %>%
  filter(!is.na(FAM_COMP)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=FAM_COMP))

els %>%
  filter(!is.na(FAM_COMP)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=FAM_COMP))

els %>%
  filter(!is.na(FAM_COMP), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=FAM_COMP), position = "fill")
```


- parents' highest level of education: BYPARED
BYPARED
   1 = "Did not finish high school"
   2 = "Graduated from high school or GED"
   3 = "Attended 2-year school, no degree"
   4 = "Graduated from 2-year school"
   5 = "Attended college, no 4-year degree"
   6 = "Graduated from college"
   7 = "Completed Master's degree or equivalent"
   8 = "Completed PhD, MD, other advanced degree"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
   -9 = "Missing"
```{r}
els <- els %>% mutate(
  BYPARED = na_if(BYPARED, BYPARED<0),
  PT_ED = case_when(
    BYPARED == 1 ~ "No HS",
    BYPARED == 2 ~ "GED",
    BYPARED == 3 | BYPARED == 5 ~ "Attended PS",
    BYPARED == 4 ~ "Associates",
    BYPARED == 6 ~ "Bachelors",
    BYPARED == 7 | BYPARED == 8 ~ "Advanced Degree"
  )
)
els$PT_ED = factor(els$PT_ED, levels=c("No HS", "GED", "Attended PS", "Associates", "Bachelors", "Advanced Degree"))

els %>%
  filter(!is.na(PT_ED)) %>%
  ggplot() +
  geom_bar(aes(x=PT_ED))

              
els %>%
  filter(!is.na(PT_ED)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=PT_ED))

els %>%
  filter(!is.na(PT_ED)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=PT_ED))

els %>%
  filter(!is.na(PT_ED), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=PT_ED), position = "fill")
```


- parents' English fluency


- socioeconomic status: BYSES1, BYSES1QU, BYSES2, BYSES2QU
```{r}

els <- els %>% mutate(
  SESQU = ifelse(BYSES1QU < 0, NA, BYSES1QU),
  SESQU=factor(SESQU),
  SES = ifelse(BYSES1 <= -2, NA, BYSES1)
)

hist(els$BYSES1)


els %>%
  filter(!is.na(SESQU)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=SESQU))

els %>%
  filter(!is.na(SESQU)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=SESQU))


els %>%
  filter(!is.na(SES), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=SES))
```

- family income: BYINCOME
BYINCOME
   1 = "None"
   2 = "$1,000 or less"
   3 = "$1,001-$5,000"
   4 = "$5,001-$10,000"
   5 = "$10,001-$15,000"
   6 = "$15,001-$20,000"
   7 = "$20,001-$25,000"
   8 = "$25,001-$35,000"
   9 = "$35,001-$50,000"
   10 = "$50,001-$75,000"
   11 = "$75,001-$100,000"
   12 = "$100,001-$200,000"
   13 = "$200,001 or more"
```{r}
# els <- els %>% mutate(
#   PT_ED = na_if(BYPARED, BYPARED<0),
#   PT_ED = case_when(
#     BYPARED == 1 ~ "No HS",
#     BYPARED == 2 ~ "GED",
#     BYPARED == 3 | BYPARED == 5 ~ "Attended PS",
#     BYPARED == 4 ~ "Associates",
#     BYPARED == 6 ~ "Bachelors",
#     BYPARED == 7 | BYPARED == 8 ~ "Advanced Degree"
#   )
# )
# els$PT_ED = factor(els$PT_ED, levels=c("No HS", "GED", "Attended PS", "Associates", "Bachelors", "Advanced Degree"))

hist(els$BYINCOME)

els %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=factor(BYINCOME)))

els %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=factor(BYINCOME)))

els %>%
  filter(!is.na(BYINCOME), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=factor(BYINCOME)), position = "fill")

```

<!-- - how far in school parents wants student to go: BYPARASP -->
<!-- BYPARASP -->
<!--    1 = "Less than high school graduation" -->
<!--    2 = "High school graduation or GED only" -->
<!--    3 = "Attend or complete 2-year college/school" -->
<!--    4 = "Attend college, 4-year degree incomplete" -->
<!--    5 = "Graduate from college" -->
<!--    6 = "Obtain Master's degree or equivalent" -->
<!--    7 = "Obtain PhD, MD, or other advanced degree" -->
<!--    -4 = "Nonrespondent" -->

- has a computer at home: BYS84C
```{r}
# els <- els %>% mutate(
#   PT_ED = na_if(BYPARED, BYPARED<0),
#   PT_ED = case_when(
#     BYPARED == 1 ~ "No HS",
#     BYPARED == 2 ~ "GED",
#     BYPARED == 3 | BYPARED == 5 ~ "Attended PS",
#     BYPARED == 4 ~ "Associates",
#     BYPARED == 6 ~ "Bachelors",
#     BYPARED == 7 | BYPARED == 8 ~ "Advanced Degree"
#   )
# )
# els$PT_ED = factor(els$PT_ED, levels=c("No HS", "GED", "Attended PS", "Associates", "Bachelors", "Advanced Degree"))

hist(els$BYS84C)

els %>%
  filter(BYS84C>=0) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=factor(BYS84C)))

els %>%
  filter(BYS84C>=0) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=factor(BYS84C)))

els %>%
  filter(BYS84C>=0, !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=factor(BYS84C)), position = "fill")

```

- has own room: BYS84I
- has Internet access: BYS84D
- parents expect success in school: BYS27I

# Educational (for the school the student attends)
- school type (private/public): BYSCTRL
- school urbanicity: BYURBAN
- dropout prevention program offered: F1A23
- programs for pregnant girls/teenage mothers offered: F1A21F
- individual/family psychological counseling offered: F1A21E
- school-sponsored community service offered: F1A21D
- vocational counseling/services/programs offered: F1A21A
- learning hindered by poor condition of buildings, poor heating/air/light, poor science labs, poor fine arts facilities, poor lack of space, library, lack of texts/supplies, too few computers, lack of multi-media, lack of discipline/safety, poor voc/tech equipment/facilities

BYA50A         14277-14279   Learning hindered by poor condition of buildings
BYA50B         14279-14281   Learning hindered by poor heating/air/light
BYA50C         14281-14283   Learning hindered by poor science labs
BYA50D         14283-14285   Learning hindered by poor fine arts facilities
BYA50E         14285-14287   Learning hindered by lack of space
BYA50F         14287-14289   Learning hindered by poor library
BYA50G         14289-14291   Learning hindered by lack of texts/supplies
BYA50H         14291-14293   Learning hindered by too few computers
BYA50I         14293-14295   Learning hindered by lack of multi-media
BYA50J         14295-14297   Learning hindered by lack of discipline/safety
BYA50K         14297-14299   Learning hindered by poor voc/tech equipment/facilities


- teacher access to technology

BYA41A         14131-14133   Teachers have access to cable TV
BYA41B         14133-14135   Teachers have access to closed-circuit TV
BYA41C         14135-14137   Teachers have access to videodisc player/VCR/DVD
BYA41D         14137-14139   Teachers have access to video camera
BYA41E         14139-14141   Teachers have access to video production studio
BYA41F         14141-14143   Teachers have access to satellite TV hook-up
BYA41G         14143-14145   Teachers have access to videoconferencing equipment
BYA41H         14145-14147   Teachers have access to digital camera
BYA41I         14147-14149   Teachers have access to scanner
BYA41J         14149-14151   Teachers have access to LCD panel
BYA41K         14151-14153   Teachers have access to laptop computer
BYA41L         14153-14155   Teachers have access to Internet
BYA41M         14155-14157   Teachers have access to computer printer

- highest salary paid to full-time teachers: BYA26B
- % full-time teachers that are certified: BYA24A
- % full-time teachers that teach out of field: BYA25A
- percent of 10th graders with free lunch: BY10FLP



```{r}

```

