---
title: "Data Exploration"
author: "Christina Chen"
date: "2/22/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)

# data file too large to push to git - split into two halfs in split_data.R
student_data1 = read.csv("els_student_first_half.csv")
student_data2 = read.csv("els_student_second_half.csv")
els = rbind(student_data1, student_data2)
els_original=els
```

# Measures of Success
- standardized test quartile: BYTXCQU 
- standardized test composite: BYTXCSTD
```{r}
# hist(student_data$BYTXCQU)
hist(els$BYTXCSTD)
# some scores below 0 -> should remove

```

- employment income 7 years after graduation (2011): F3ERN2011
- Socioeconomic status in 2011 F3SESQU
```{r}
sum(els$F3ERN2011==0)
hist(log(els$F3ERN2011))

els <- els %>%
  mutate(F3ERN2011_ORIGINAL = F3ERN2011,
         F3ERN2011 = ifelse(F3ERN2011 < 0, NA, F3ERN2011),
         F3ERN2011 = log(F3ERN2011+0.1))
```

- highest level of education earned 8 years after graduation (2012): F3ATTAINMENT
F3ATTAINMENT
   1 = "No HS credential, no PS attendance"
   2 = "HS credential, no PS attendance"
   3 = "Some PS attendance, no PS credential"
   4 = "Undergraduate certificate"
   5 = "Associates degree"
   6 = "Bachelors degree"
   7 = "Post-Baccalaureate certificate"
   8 = "Master's degree/Post-Master's certificate"
   10 = "Doctoral degree"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
```{r}
els <- els %>% mutate(
  ED_ATNMT = case_when(
    F3ATTAINMENT == 1 ~ "No GED",
    F3ATTAINMENT == 2 ~ "GED",
    F3ATTAINMENT ==3 | F3ATTAINMENT ==4 | F3ATTAINMENT ==5 ~ "Associates/Some Undergrad",
    F3ATTAINMENT ==6 ~ "Bachelors",
    F3ATTAINMENT ==7 | F3ATTAINMENT ==8~ "Master's/Post-Bacc. Cert.",
    F3ATTAINMENT ==10 ~ "PhD"
  )
)
els$ED_ATNMT = factor(els$ED_ATNMT, levels=c( "No GED", "GED", "Associates/Some Undergrad", "Bachelors", "Master's/Post-Bacc. Cert.", "PhD"))

els %>%
  filter(!is.na(ED_ATNMT)) %>%
  ggplot() +
    geom_bar(aes(x=ED_ATNMT)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5)) +
  labs(x="Education Attained", y="Count")
```

## relationships between outcomes
```{r}
els %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=F3ERN2011))

els %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=BYTXCSTD))

els %>%
  filter(F3ERN2011!=0) %>%
  ggplot() +
  geom_point(aes(x=F3ERN2011, y =BYTXCSTD), alpha=0.05)
```

<!-- - HS GPA: F1RGPP2 -->

<!-- - SAT: TXEESATC -->

<!-- - ACT: TXEEACTC -->

# Familial
- race: BYRACE
BYRACE
   1 = "Amer. Indian/Alaska Native, non-Hispanic"
   2 = "Asian, Hawaii/Pac. Islander,non-Hispanic"
   3 = "Black or African American, non-Hispanic"
   4 = "Hispanic, no race specified"
   5 = "Hispanic, race specified"
   6 = "More than one race, non-Hispanic"
   7 = "White, non-Hispanic"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
```{r}
# AI/AN = American Indian/Alaska Native
# API = Asian, Hawaii/Pacific Islander
els <- els %>% mutate(
  RACE = case_when(
    BYRACE == 1 ~ "AI/AN",
    BYRACE ==2 ~ "API",
    BYRACE ==3 ~ "Black",
    BYRACE ==4 | BYRACE ==5 ~ "Hispanic",
    BYRACE ==6 ~ "Mixed",
    BYRACE ==7 ~ "White"
  )
)
els$RACE = factor(els$RACE, levels=c( "AI/AN", "Mixed", "API", "Black", "Hispanic","White"))

els %>%
  filter(!is.na(RACE)) %>%
  ggplot() +
  geom_bar(aes(x=RACE))

els$RACE = factor(els$RACE, levels=c("Black", "Hispanic", "AI/AN", "Mixed", "API", "White"))
              
els %>%
  filter(!is.na(RACE)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=RACE))

els %>%
  filter(!is.na(RACE)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=RACE))

els %>%
  filter(!is.na(RACE), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=RACE, fill=ED_ATNMT), position = "fill")

sum(is.na(els$RACE))
```

- family composition: BYFCOMP

BYFCOMP
   1 = "Mother and father"
   2 = "Mother and male guardian"
   3 = "Father and female guardian"
   4 = "Two guardians"
   5 = "Mother only"
   6 = "Father only"
   7 = "Female guardian only"
   8 = "Male guardian only"
   9 = "Lives with student less than half time"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
   -9 = "Missing"
```{r}
els <- els %>% mutate(
  BYFCOMP = na_if(BYFCOMP, BYFCOMP<0),
  FAM_COMP = case_when(
    BYFCOMP == 1 | BYFCOMP == 2 | BYFCOMP == 3 | BYFCOMP == 4 ~ "Two Parents",
    BYFCOMP ==5 | BYFCOMP == 6 | BYFCOMP ==7 | BYFCOMP == 8 ~ "Single Parent",
    BYFCOMP == 9 ~ "Absent Parent"
  )
)
# els$RACE = factor(els$RACE, levels=c( "Amer. Indian/Alaska Native", "Mixed", "Asian, Hawaii/Pac. Islander", "Black", "Hispanic","White"))


els %>%
  filter(!is.na(FAM_COMP)) %>%
  ggplot() +
  geom_bar(aes(x=FAM_COMP))

# els$RACE = factor(els$RACE, levels=c("Black", "Hispanic", "Amer. Indian/Alaska Native", "Mixed", "Asian, Hawaii/Pac. Islander", "White"))
              
els %>%
  filter(!is.na(FAM_COMP)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=FAM_COMP))

els %>%
  filter(!is.na(FAM_COMP)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=FAM_COMP))

els %>%
  filter(!is.na(FAM_COMP), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=FAM_COMP), position = "fill")
```


- parents' highest level of education: BYPARED
BYPARED
   1 = "Did not finish high school"
   2 = "Graduated from high school or GED"
   3 = "Attended 2-year school, no degree"
   4 = "Graduated from 2-year school"
   5 = "Attended college, no 4-year degree"
   6 = "Graduated from college"
   7 = "Completed Master's degree or equivalent"
   8 = "Completed PhD, MD, other advanced degree"
   -4 = "Nonrespondent"
   -8 = "Survey component legitimate skip/NA"
   -9 = "Missing"
```{r}
els <- els %>% mutate(
  BYPARED = na_if(BYPARED, BYPARED<0),
  PT_ED = case_when(
    BYPARED == 1 ~ "No HS",
    BYPARED == 2 ~ "GED",
    BYPARED == 3 | BYPARED == 5 ~ "Attended PS",
    BYPARED == 4 ~ "Associates",
    BYPARED == 6 ~ "Bachelors",
    BYPARED == 7 | BYPARED == 8 ~ "Advanced Degree"
  )
)
els$PT_ED = factor(els$PT_ED, levels=c("No HS", "GED", "Attended PS", "Associates", "Bachelors", "Advanced Degree"))

els %>%
  filter(!is.na(PT_ED)) %>%
  ggplot() +
  geom_bar(aes(x=PT_ED))

              
els %>%
  filter(!is.na(PT_ED)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=PT_ED))


els %>%
  filter(!is.na(PT_ED)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=PT_ED))

els %>%
  filter(!is.na(PT_ED), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=PT_ED), position = "fill")

```


- parents' English fluency


- socioeconomic status: BYSES1, BYSES1QU, BYSES2, BYSES2QU
```{r}

els <- els %>% mutate(
  SESQU = ifelse(BYSES1QU < 0, NA, BYSES1QU),
  SESQU=factor(SESQU),
  SES = ifelse(BYSES1 <= -2, NA, BYSES1)
)

hist(els$BYSES1)


els %>%
  filter(!is.na(SESQU)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=SESQU))

els %>%
  filter(!is.na(SESQU)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=SESQU))


els %>%
  filter(!is.na(SES), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=SES))
```

- family income: BYINCOME
BYINCOME
   1 = "None"
   2 = "$1,000 or less"
   3 = "$1,001-$5,000"
   4 = "$5,001-$10,000"
   5 = "$10,001-$15,000"
   6 = "$15,001-$20,000"
   7 = "$20,001-$25,000"
   8 = "$25,001-$35,000"
   9 = "$35,001-$50,000"
   10 = "$50,001-$75,000"
   11 = "$75,001-$100,000"
   12 = "$100,001-$200,000"
   13 = "$200,001 or more"
```{r}
# els <- els %>% mutate(
#   PT_ED = na_if(BYPARED, BYPARED<0),
#   PT_ED = case_when(
#     BYPARED == 1 ~ "No HS",
#     BYPARED == 2 ~ "GED",
#     BYPARED == 3 | BYPARED == 5 ~ "Attended PS",
#     BYPARED == 4 ~ "Associates",
#     BYPARED == 6 ~ "Bachelors",
#     BYPARED == 7 | BYPARED == 8 ~ "Advanced Degree"
#   )
# )
# els$PT_ED = factor(els$PT_ED, levels=c("No HS", "GED", "Attended PS", "Associates", "Bachelors", "Advanced Degree"))

hist(els$BYINCOME)

# els %>%
#   ggplot() +
#   geom_bar(aes(x=BYINCOME, fill=factor(SESQU)), position="fill")

els %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=factor(BYINCOME)))

els %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=factor(BYINCOME)))

els %>%
  filter(!is.na(BYINCOME), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=factor(BYINCOME)), position = "fill")

els <- els %>% mutate(FAM_INCOME = factor(BYINCOME))
```

- how far in school parents wants student to go: BYPARASP
BYPARASP
   1 = "Less than high school graduation"
   2 = "High school graduation or GED only"
   3 = "Attend or complete 2-year college/school"
   4 = "Attend college, 4-year degree incomplete"
   5 = "Graduate from college"
   6 = "Obtain Master's degree or equivalent"
   7 = "Obtain PhD, MD, or other advanced degree"
   -4 = "Nonrespondent"
```{r}
els <- els %>% mutate(
  PT_PUSH = na_if(BYPARASP, BYPARASP<0),
  PT_PUSH = case_when(
    PT_PUSH == 1 | PT_PUSH ==2 ~ "Some HS/GED",
    PT_PUSH == 3 | BYPARED == 4 ~ "Any college (2 or 4 year)",
    PT_PUSH == 5 ~ "Graduate from college",
    PT_PUSH == 6 ~ "Master's",
    PT_PUSH == 7 ~ "Advanced Degree"
  )
)
els$PT_PUSH = factor(els$PT_PUSH, levels=c("Some HS/GED", "Any college (2 or 4 year)", "Graduate from college", "Master's", "Advanced Degree"))

els %>%
  ggplot() +
  geom_bar(aes(x=PT_PUSH))
# hist(els$BYINCOME)

els %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=PT_PUSH))

els %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=PT_PUSH))

els %>%
  filter(!is.na(PT_PUSH), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=PT_PUSH), position = "fill")

els %>%
  filter(!is.na(PT_ED), !is.na(PT_PUSH)) %>%
  ggplot() +
  geom_bar(aes(x=PT_PUSH, fill=PT_ED), position = "fill")

```

- has a computer at home: BYS84C
```{r}
els <- els %>% mutate(
  COMP_HOME = ifelse(BYS84C < 0, NA, BYS84C),
  COMP_HOME=factor(COMP_HOME)
)

els %>%
  filter(!is.na(COMP_HOME)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=COMP_HOME))

els %>%
  filter(!is.na(COMP_HOME)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=COMP_HOME))

els %>%
  filter(!is.na(COMP_HOME), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=COMP_HOME), position = "fill")

```

- has own room: BYS84I
```{r}
els <- els %>% mutate(
  OWN_ROOM = ifelse(BYS84I < 0, NA, BYS84I),
  OWN_ROOM=factor(OWN_ROOM)
)

els %>%
  filter(!is.na(OWN_ROOM)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=OWN_ROOM))

els %>%
  filter(!is.na(OWN_ROOM)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=OWN_ROOM))

els %>%
  filter(!is.na(OWN_ROOM), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=OWN_ROOM), position = "fill")

```

- has Internet access: BYS84D
```{r}
els <- els %>% mutate(
  HAS_INTERNET = ifelse(BYS84D < 0, NA, BYS84D),
  HAS_INTERNET=factor(HAS_INTERNET)
)

els %>%
  filter(!is.na(HAS_INTERNET)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=HAS_INTERNET))

els %>%
  filter(!is.na(HAS_INTERNET)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=HAS_INTERNET))

els %>%
  filter(!is.na(HAS_INTERNET), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=HAS_INTERNET), position = "fill")

```
- parents expect success in school: BYS27I
BYS27I
   1 = "Strongly agree"
   2 = "Agree"
   3 = "Disagree"
   4 = "Strongly disagree"
   -4 = "Nonrespondent"
   -6 = "Multiple response"
   -7 = "Not administered; abbreviated interview or breakoff"
   -8 = "Survey component legitimate skip/NA"
   -9 = "Missing"
```{r}
els <- els %>% mutate(
  PT_EXPT_SCSS = ifelse(BYS27I < 0, NA, BYS27I),
  PT_EXPT_SCSS=case_when(
    PT_EXPT_SCSS == 1 ~ "Strongly Agree",
    PT_EXPT_SCSS == 2 ~ "Agree",
    PT_EXPT_SCSS == 3 ~ "Disagree",
    PT_EXPT_SCSS == 4 ~ "Strongly Disagree"
  )
)

els %>%
  filter(!is.na(PT_EXPT_SCSS)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=PT_EXPT_SCSS))

els %>%
  filter(!is.na(PT_EXPT_SCSS)) %>%
  ggplot() +
  geom_boxplot(aes(y=log(F3ERN2011), x=PT_EXPT_SCSS))

els %>%
  filter(!is.na(PT_EXPT_SCSS), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=PT_EXPT_SCSS), position = "fill")

```

# Educational (for the school the student attends)
- school type (private/public): BYSCTRL
BYSCTRL
   1 = "Public"
   2 = "Catholic"
   3 = "Other private"
```{r}
els <- els %>% mutate(
  SCHOOL_TYPE=case_when(
    BYSCTRL == 1 ~ "Public",
    BYSCTRL == 2 ~ "Catholic",
    BYSCTRL == 3 ~ "Private"
  )
)

els %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=SCHOOL_TYPE))

els %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=SCHOOL_TYPE))

els %>%
  filter(!is.na(SCHOOL_TYPE), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=SCHOOL_TYPE), position = "fill")

```
- school urbanicity: BYURBAN
BYURBAN
   1 = "Urban"
   2 = "Suburban"
   3 = "Rural"
```{r}
els <- els %>% mutate(
  SCHOOL_URBAN=case_when(
    BYURBAN == 1 ~ "Urban",
    BYURBAN == 2 ~ "Suburban",
    BYURBAN == 3 ~ "Rural"
  )
)

els %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=SCHOOL_URBAN))

els %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=SCHOOL_URBAN))

els %>%
  filter(!is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=SCHOOL_URBAN), position = "fill")

```
- dropout prevention program offered: F1A23
```{r}
els <- els %>% mutate(
  DROPOUT_PREV_PGM = ifelse(F1A23 < 0, NA, F1A23),
  DROPOUT_PREV_PGM=factor(DROPOUT_PREV_PGM)
)


els %>%
  filter(!is.na(DROPOUT_PREV_PGM)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=DROPOUT_PREV_PGM))

els %>%
  filter(!is.na(DROPOUT_PREV_PGM)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=DROPOUT_PREV_PGM))

els %>%
  filter(!is.na(DROPOUT_PREV_PGM), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=DROPOUT_PREV_PGM), position = "fill")

els %>%
  ggplot() +
  geom_bar(aes(fill=DROPOUT_PREV_PGM, x=FREE_LUNCH_PERC), position="fill")

```
- programs for pregnant girls/teenage mothers offered: F1A21F
```{r}
els <- els %>% mutate(
  PRG_PGM = ifelse(F1A21F < 0, NA, F1A21F),
  PRG_PGM=factor(PRG_PGM)
)


els %>%
  filter(!is.na(PRG_PGM)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=PRG_PGM))

els %>%
  filter(!is.na(PRG_PGM)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=PRG_PGM))

els %>%
  filter(!is.na(PRG_PGM), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=PRG_PGM), position = "fill") +
  facet_grid(.~SESQU)

```
- individual/family psychological counseling offered: F1A21E
```{r}
els <- els %>% mutate(
  PSYCH_COUNSELING = ifelse(F1A21E < 0, NA, F1A21E),
  PSYCH_COUNSELING=factor(PSYCH_COUNSELING)
)


els %>%
  filter(!is.na(PSYCH_COUNSELING)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=PSYCH_COUNSELING))

els %>%
  filter(!is.na(PSYCH_COUNSELING)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=PSYCH_COUNSELING))

els %>%
  filter(!is.na(PSYCH_COUNSELING), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=PSYCH_COUNSELING), position = "fill")

```
- school-sponsored community service offered: F1A21D
```{r}
els <- els %>% mutate(
  COMMTY_SERVICE = ifelse(F1A21D < 0, NA, F1A21D),
  COMMTY_SERVICE=factor(COMMTY_SERVICE)
)


els %>%
  filter(!is.na(COMMTY_SERVICE)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=COMMTY_SERVICE))

els %>%
  filter(!is.na(COMMTY_SERVICE)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=COMMTY_SERVICE))

els %>%
  filter(!is.na(COMMTY_SERVICE), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=COMMTY_SERVICE), position = "fill")

```
- vocational counseling/services/programs offered: F1A21A
```{r}
els <- els %>% mutate(
  VOCATIONAL_PGM = ifelse(F1A21A < 0, NA, F1A21A),
  VOCATIONAL_PGM=factor(VOCATIONAL_PGM)
)


els %>%
  filter(!is.na(VOCATIONAL_PGM)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=VOCATIONAL_PGM))

els %>%
  filter(!is.na(VOCATIONAL_PGM)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=VOCATIONAL_PGM))

els %>%
  filter(!is.na(VOCATIONAL_PGM), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=VOCATIONAL_PGM), position = "fill")

```
- learning hindered by poor condition of buildings, poor heating/air/light, poor science labs, poor fine arts facilities, poor lack of space, library, lack of texts/supplies, too few computers, lack of multi-media, lack of discipline/safety, poor voc/tech equipment/facilities

BYA50A         14277-14279   Learning hindered by poor condition of buildings
BYA50B         14279-14281   Learning hindered by poor heating/air/light
BYA50C         14281-14283   Learning hindered by poor science labs
BYA50D         14283-14285   Learning hindered by poor fine arts facilities
BYA50E         14285-14287   Learning hindered by lack of space
BYA50F         14287-14289   Learning hindered by poor library

BYA50G         14289-14291   Learning hindered by lack of texts/supplies
BYA50H         14291-14293   Learning hindered by too few computers
BYA50I         14293-14295   Learning hindered by lack of multi-media
BYA50J         14295-14297   Learning hindered by lack of discipline/safety
BYA50K         14297-14299   Learning hindered by poor voc/tech equipment/facilities

```{r}

els <- els %>% mutate(
  BYA50A_1 = ifelse(BYA50A<0, NA, BYA50A),
  BYA50B_1 = ifelse(BYA50B<0, NA, BYA50B),
  BYA50C_1 = ifelse(BYA50C<0, NA, BYA50C),
  BYA50D_1 = ifelse(BYA50D<0, NA, BYA50D),
  BYA50E_1 = ifelse(BYA50E<0, NA, BYA50E),
  BYA50F_1 = ifelse(BYA50F<0, NA, BYA50F),
  BYA50G_1 = ifelse(BYA50G<0, NA, BYA50G),
  BYA50H_1 = ifelse(BYA50H<0, NA, BYA50H),
  BYA50I_1 = ifelse( BYA50I<0,NA, BYA50I),
  BYA50J_1 = ifelse( BYA50J<0, NA, BYA50J),
  BYA50K_1 = ifelse(BYA50K<0, NA, BYA50K),
  LEARNING_HINDERED_CONDITIONS = BYA50A_1 + BYA50B_1,
  LEARNING_HINDERED_POOR_FACILITIES = BYA50C_1+BYA50D_1+BYA50F_1,
  LEARNING_HINDERED_LACK_SPACE = BYA50E_1,
  LEARNING_HINDERED_LACK_SUPPLIES = 2*BYA50G_1+ BYA50H_1+BYA50I+BYA50K_1,
  # LEARNING_HINDERED = LEARNING_HINDERED_CONDITIONS + LEARNING_HINDERED_LACK_SUPPLIES,
  # LEARNING_HINDERED_QU = case_when(
  #   LEARNING_HINDERED <=15 ~ 1,
  #   LEARNING_HINDERED <=20 ~ 2,
  #   LEARNING_HINDERED <=27 ~ 3,
  #   LEARNING_HINDERED <=49 ~ 4
  # ),
  # LEARNING_HINDERED_QU = factor(LEARNING_HINDERED_QU)
)

els %>%
  filter(!is.na(LEARNING_HINDERED_CONDITIONS)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=LEARNING_HINDERED_CONDITIONS))

els %>%
  filter(!is.na(LEARNING_HINDERED_CONDITIONS)) %>%
  ggplot() +
  geom_point(aes(y=F3ERN2011, x=LEARNING_HINDERED_CONDITIONS))

els %>%
  filter(!is.na(LEARNING_HINDERED_CONDITIONS), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=LEARNING_HINDERED_CONDITIONS))

els %>%
  filter(!is.na(LEARNING_HINDERED_LACK_SUPPLIES)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=LEARNING_HINDERED_LACK_SUPPLIES))

els %>%
  filter(!is.na(LEARNING_HINDERED_LACK_SUPPLIES)) %>%
  ggplot() +
  geom_point(aes(y=F3ERN2011, x=LEARNING_HINDERED_LACK_SUPPLIES))

els %>%
  filter(!is.na(LEARNING_HINDERED_LACK_SUPPLIES), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=LEARNING_HINDERED_LACK_SUPPLIES))

els %>%
  filter(!is.na(LEARNING_HINDERED)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=LEARNING_HINDERED_QU))

els %>%
  filter(!is.na(LEARNING_HINDERED)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=LEARNING_HINDERED_QU))

els %>%
  filter(!is.na(LEARNING_HINDERED), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=LEARNING_HINDERED))



els %>%
  filter(!is.na(BYA50A_1)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=factor(BYA50A_1)))

els %>%
  filter(!is.na(BYA50A_1)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=factor(BYA50A_1)))

els %>%
  filter(!is.na(BYA50A_1), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=factor(BYA50A_1)), position="fill")
# els %>%
#   filter(!is.na(LEARNING_HINDERED_CONDITIONS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=BYTXCSTD, x=LEARNING_HINDERED_CONDITIONS))
# 
# els %>%
#   filter(!is.na(LEARNING_HINDERED_CONDITIONS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=F3ERN2011, x=LEARNING_HINDERED_CONDITIONS))
# 
# els %>%
#   filter(!is.na(LEARNING_HINDERED_CONDITIONS), !is.na(ED_ATNMT)) %>%
#   ggplot() +
#   geom_bar(aes(x=ED_ATNMT, fill=LEARNING_HINDERED_CONDITIONS), position = "fill")
# 
# els %>%
#   filter(!is.na(LEARNING_HINDERED_LACK_SUPPLIES)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=BYTXCSTD, x=LEARNING_HINDERED_LACK_SUPPLIES))
# 
# els %>%
#   filter(!is.na(LEARNING_HINDERED_LACK_SUPPLIES)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=F3ERN2011, x=LEARNING_HINDERED_LACK_SUPPLIES))
# 
# els %>%
#   filter(!is.na(LEARNING_HINDERED_LACK_SUPPLIES), !is.na(ED_ATNMT)) %>%
#   ggplot() +
#   geom_bar(aes(x=ED_ATNMT, fill=LEARNING_HINDERED_LACK_SUPPLIES), position = "fill")

```


- teacher access to technology

BYA41A         14131-14133   Teachers have access to cable TV
BYA41B         14133-14135   Teachers have access to closed-circuit TV
BYA41C         14135-14137   Teachers have access to videodisc player/VCR/DVD
BYA41D         14137-14139   Teachers have access to video camera
BYA41E         14139-14141   Teachers have access to video production studio
BYA41F         14141-14143   Teachers have access to satellite TV hook-up
BYA41G         14143-14145   Teachers have access to videoconferencing equipment
BYA41H         14145-14147   Teachers have access to digital camera
BYA41I         14147-14149   Teachers have access to scanner
BYA41J         14149-14151   Teachers have access to LCD panel
BYA41K         14151-14153   Teachers have access to laptop computer
BYA41L         14153-14155   Teachers have access to Internet
BYA41M         14155-14157   Teachers have access to computer printer


BYA42A         14157-14159   Teachers use computers as instructional tools
BYA42B         14159-14161   Teachers use computers to plan lessons
BYA42C         14161-14163   Teachers use computers for professional development courses
BYA42D         14163-14165   Teachers use computers to communicate with colleagues
BYA42E         14165-14167   Teachers use computers to access best practices
BYA42F         14167-14169   Teachers use computers to communicate with parents
BYA42G         14169-14171   Teachers use computers to post homework
BYA42H         14171-14173   Teachers/staff use computers to communicate with each other
BYA42I         14173-14175   Teachers use computers to teach job skills

BYA43A         14185-14187   Teacher training on use of new software
BYA43B         14187-14189   Teacher training on use of Internet
BYA43C         14189-14191   Teacher training on using computers to teach skills
BYA43D         14191-14193   Teacher training on integrating computer into class
BYA43E         14193-14195   Teacher training on basic computer literacy
```{r}

els <- els %>% mutate(
  BYA41A_1 = ifelse( BYA41A<0, NA, BYA41A),
  BYA41B_1 = ifelse( BYA41B<0, NA, BYA41B),
  BYA41C_1 = ifelse( BYA41C<0, NA, BYA41C),
  BYA41D_1 = ifelse( BYA41D<0, NA, BYA41D),
  BYA41E_1 = ifelse( BYA41E<0, NA, BYA41E),
  BYA41F_1 = ifelse( BYA41F<0, NA, BYA41F),
  BYA41G_1 = ifelse( BYA41G<0, NA, BYA41G),
  BYA41H_1 = ifelse( BYA41H<0, NA, BYA41H),
  BYA41I_1 = ifelse( BYA41I<0, NA, BYA41I),
  BYA41J_1 = ifelse( BYA41J<0, NA, BYA41J),
  BYA41K_1 = ifelse( BYA41K<0, NA, BYA41K),
  BYA41L_1 = ifelse( BYA41L<0, NA, BYA41L),
  BYA41M_1 = ifelse(BYA41M<0, NA, BYA41M),
  TEACHER_BASIC_TECH_ACCESS = factor(BYA41K_1 & BYA41L_1 & BYA41M_1),
  TEACHER_TECH_ACCESS = BYA41A_1+ BYA41B_1+BYA41C_1+BYA41D_1+BYA41E_1+BYA41F_1+BYA41G_1+BYA41H_1+BYA41I_1+BYA41J_1+2*BYA41K_1+2*BYA41L_1+ BYA41M_1
)

hist(els$TEACHER_TECH_ACCESS)

els %>%
  filter(!is.na(TEACHER_TECH_ACCESS)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=TEACHER_TECH_ACCESS))

els %>%
  filter(!is.na(TEACHER_TECH_ACCESS)) %>%
  ggplot() +
  geom_point(aes(y=F3ERN2011, x=TEACHER_TECH_ACCESS))

els %>%
  filter(!is.na(TEACHER_TECH_ACCESS), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=TEACHER_TECH_ACCESS))



# els %>%
#   filter(!is.na(TEACHER_BASIC_TECH_ACCESS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=BYTXCSTD, x=TEACHER_BASIC_TECH_ACCESS))
# 
# els %>%
#   filter(!is.na(TEACHER_BASIC_TECH_ACCESS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=F3ERN2011, x=TEACHER_BASIC_TECH_ACCESS))
# 
# els %>%
#   filter(!is.na(TEACHER_BASIC_TECH_ACCESS), !is.na(ED_ATNMT)) %>%
#   ggplot() +
#   geom_bar(aes(x=ED_ATNMT, fill=TEACHER_BASIC_TECH_ACCESS), position="fill")


# els %>%
#   filter(!is.na(TEACHER_TECH_ACCESS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=BYTXCSTD, x=TEACHER_TECH_ACCESS))
# 
# els %>%
#   filter(!is.na(TEACHER_TECH_ACCESS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=F3ERN2011, x=TEACHER_TECH_ACCESS))
# 
# els %>%
#   filter(!is.na(TEACHER_TECH_ACCESS), !is.na(ED_ATNMT)) %>%
#   ggplot() +
#   geom_bar(aes(x=ED_ATNMT, fill=TEACHER_TECH_ACCESS), position = "fill")
# 
# 
# 
# els %>%
#   filter(!is.na(BYA41B_1)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=BYTXCSTD, x=factor(BYA41B_1)))
# 
# els %>%
#   filter(!is.na(BYA41B_1)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=F3ERN2011, x=factor(BYA41B_1)))
# 
# els %>%
#   filter(!is.na(BYA41B_1), !is.na(ED_ATNMT)) %>%
#   ggplot() +
#   geom_bar(aes(x=ED_ATNMT, fill=factor(BYA41B_1)), position = "fill")

```
- highest salary paid to full-time teachers: BYA26B
- lowest salary paid to full-time teachers: BYA26A
```{r}
els <- els %>% mutate(
  HIGHEST_TEACHER_SALARY = ifelse(BYA26B < 0, NA, BYA26B),
  LOWEST_TEACHER_SALARY = ifelse(BYA26A < 0, NA, BYA26A)
)

els %>%
  ggplot() +
  geom_point(aes(x=HIGHEST_TEACHER_SALARY,y=LOWEST_TEACHER_SALARY))


els %>%
  filter(!is.na(HIGHEST_TEACHER_SALARY)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=HIGHEST_TEACHER_SALARY), alpha=0.1) +
  facet_grid(.~SCHOOL_TYPE)

els %>%
  filter(!is.na(HIGHEST_TEACHER_SALARY)) %>%
  ggplot() +
  geom_point(aes(y=F3ERN2011, x=HIGHEST_TEACHER_SALARY), alpha=0.1) +
  facet_grid(.~SCHOOL_TYPE)

els %>%
  filter(!is.na(HIGHEST_TEACHER_SALARY), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=HIGHEST_TEACHER_SALARY)) +
  facet_grid(.~SCHOOL_TYPE)

```

- % full-time teachers that are certified: BYA24A/ F1A34A
```{r}
els <- els %>% mutate(
  F1A34A_QU = case_when(
    F1A34A <= 33 ~ 1,
    F1A34A <= 67 ~ 2,
    F1A34A <= 100 ~ 3
  ),
  F1A34A_QU = factor(F1A34A_QU)
)
hist(els$BYA24A - els$F1A34A)

els %>%
  filter(!is.na(F1A34A)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=F1A34A), alpha=0.1)

els %>%
  filter(!is.na(F1A34A)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=F1A34A_QU)) +
  facet_grid(.~SESQU)

els %>%
  filter(!is.na(F1A34A), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=F1A34A))

```
- % full-time teachers that teach out of field: BYA25A
```{r}
# els <- els %>% mutate(
#   HIGHEST_TEACHER_SALARY = ifelse(BYA24A < 0, NA, BYA24A),
#   LOWEST_TEACHER_SALARY = ifelse(BYA26A < 0, NA, BYA26A)
# )
hist(els$BYA25A)

els %>%
  filter(!is.na(BYA25A)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=BYA25A), alpha=0.1)

els %>%
  filter(!is.na(BYA25A)) %>%
  ggplot() +
  geom_point(aes(y=F3ERN2011, x=BYA25A), alpha=0.1)

els %>%
  filter(!is.na(BYA25A), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=BYA25A))

```
- percent of 10th graders with free lunch: BY10FLP
F1SCFLP
F1SCFLP
   1 = "0-5 percent"
   2 = "6-10 percent"
   3 = "11-20 percent"
   4 = "21-30 percent"
   5 = "31-50 percent"
   6 = "51-75 percent"
   7 = "76-100 percent"
```{r}
els <- els %>% mutate(
  FREE_LUNCH_PERC = factor(ifelse(F1SCFLP < 0, NA, F1SCFLP))
)
# hist(els$FREE_LUNCH_PERC)
els%>% ggplot() + geom_bar(aes(x=FREE_LUNCH_PERC))



els <- els %>% mutate(
  FREE_LUNCH_CATEGORIES = case_when(
    as.numeric(FREE_LUNCH_PERC) <= 1 ~ "<=5%",
    as.numeric(FREE_LUNCH_PERC) <= 4 ~ "<=30%",
    as.numeric(FREE_LUNCH_PERC) <= 5 ~ "<=50%",
    # as.numeric(FREE_LUNCH_PERC) <= 6 ~ "<=75%",
    as.numeric(FREE_LUNCH_PERC) <= 7 ~ ">50%"
  )

)
els$FREE_LUNCH_CATEGORIES = factor(els$FREE_LUNCH_CATEGORIES, levels = c("<=5%",  "<=30%", "<=50%",  ">50%"))

els %>%
  filter(!is.na(FREE_LUNCH_CATEGORIES)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=FREE_LUNCH_CATEGORIES))

els %>%
  filter(!is.na(FREE_LUNCH_CATEGORIES)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=FREE_LUNCH_CATEGORIES))

els %>%
  filter(!is.na(FREE_LUNCH_CATEGORIES), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=FREE_LUNCH_CATEGORIES), position="fill")


```

- computer access
BYA44A         14195-14197   Computers in administrative offices
BYA44B         14197-14199   Computers in teacher work rooms
BYA44C         14199-14201   Computers in classrooms
BYA44D         14201-14203   Computers in the library media center
BYA44E         14203-14205   Computers in separate computer lab
```{r}
els <- els %>% mutate(
  COMPUTER_CLASSROOMS = factor(ifelse(BYA44C < 0, NA, BYA44C)),
  COMPUTER_LABS = factor(ifelse(BYA44E < 0, NA, BYA44E))
)


els %>%
  filter(!is.na(COMPUTER_CLASSROOMS)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=COMPUTER_CLASSROOMS))
els %>%
  filter(!is.na(COMPUTER_CLASSROOMS)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=COMPUTER_CLASSROOMS))

els %>%
  filter(!is.na(COMPUTER_CLASSROOMS), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT, fill=COMPUTER_CLASSROOMS), position = "fill")



# els %>%
#   filter(!is.na(COMPUTER_LABS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=BYTXCSTD, x=COMPUTER_LABS))
# 
# els %>%
#   filter(!is.na(COMPUTER_LABS)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=F3ERN2011, x=COMPUTER_LABS))
# 
# els %>%
#   filter(!is.na(COMPUTER_LABS), !is.na(ED_ATNMT)) %>%
#   ggplot() +
#   geom_bar(aes(x=ED_ATNMT, fill=COMPUTER_LABS), position = "fill")

```

BYA40A: paid security
```{r}


els <- els %>% mutate(
  SECURITY = factor(ifelse(BYA40A < 0, NA, BYA40A)),
  FREE_LUNCH_CATEGORIES = case_when(
    as.numeric(FREE_LUNCH_PERC) <= 5 ~ "<=50%",
    as.numeric(FREE_LUNCH_PERC) > 5 ~ ">50%"
  )
)


# els %>%
#   filter(!is.na(FREE_LUNCH_CATEGORIES)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=BYTXCSTD, x=FREE_LUNCH_CATEGORIES)) +
#   facet_grid(.~ SECURITY)
# els %>%
#   filter(!is.na(FREE_LUNCH_CATEGORIES)) %>%
#   ggplot() +
#   geom_boxplot(aes(y=F3ERN2011, x=FREE_LUNCH_CATEGORIES))+
#   facet_grid(.~ SECURITY)
# 
# els %>%
#   filter(!is.na(FREE_LUNCH_CATEGORIES), !is.na(ED_ATNMT)) %>%
#   ggplot() +
#   geom_bar(aes(x=ED_ATNMT, fill=FREE_LUNCH_CATEGORIES), position = "fill")+
#   facet_grid(.~ SECURITY)

# els%>%
#   filter(!is.na(FREE_LUNCH_CATEGORIES), !is.na(ED_ATNMT)) %>%
#   ggplot()+
#   geom_point(aes(y=HIGHEST_TEACHER_SALARY, x=F3ERN2011)) +
#   facet_grid(.~ED_ATNMT)

els %>%
  filter(!is.na(F1A37D)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=F1A37D), alpha=0.1)

els %>%
  filter(!is.na(F1A37D)) %>%
  ggplot() +
  geom_point(aes(y=F3ERN2011, x=F1A37D), alpha=0.1)

els %>%
  filter(!is.na(F1A37D), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=F1A37D))


```

BYA17A         13787-13789   Agriculture/renewable resource courses offered
BYA17B         13789-13791   Business courses offered
BYA17C         13791-13793   Marketing/distribution courses offered
BYA17D         13793-13795   Health care courses offered
BYA17E         13795-13797   Public/protective service courses offered
BYA17F         13797-13799   Construction courses offered
BYA17G         13799-13801   Mechanics and repair courses offered
BYA17H         13801-13803   Precisions production courses offered
BYA17I         13803-13805   Trade/industry/transportation courses offered
BYA17J         13805-13807   Computer technology courses offered
BYA17K         13807-13809   Communication technology courses offered
BYA17L         13809-13811   Other technology courses offered
BYA17M         13811-13813   Food service and hospitality courses offered
BYA17N         13813-13815   Child care/education courses offered
BYA17O         13815-13817   Personal and other services courses offered
BYA17P         13817-13819   Other occupational courses offered
BYA17Q         13819-13821   Family/consumer sciences courses offered
BYA17R         13821-13823   Industrial arts/technology courses offered

```{r}
els <- els %>% mutate(
  BYA17A = ifelse( BYA17A<0, NA, BYA17A==1),
  BYA17B = ifelse( BYA17B<0, NA, BYA17B==1),
  BYA17C = ifelse( BYA17C<0, NA, BYA17C==1),
  BYA17D = ifelse( BYA17D<0, NA, BYA17D==1),
  BYA17E = ifelse( BYA17E<0, NA, BYA17E==1),
  BYA17F = ifelse( BYA17F<0, NA, BYA17F==1),
  BYA17G = ifelse( BYA17G<0, NA, BYA17G==1),
  BYA17H = ifelse( BYA17H<0, NA, BYA17H==1),
  BYA17I = ifelse( BYA17I<0, NA, BYA17I==1),
  BYA17J = ifelse( BYA17J<0, NA, BYA17J==1),
  BYA17K = ifelse( BYA17K<0, NA, BYA17K==1),
  BYA17L = ifelse( BYA17L<0, NA, BYA17L==1),
  BYA17M = ifelse(BYA17M<0, NA, BYA17M==1),
  BYA17N = ifelse(BYA17N<0, NA, BYA17N==1),
  BYA17O = ifelse(BYA17O<0, NA, BYA17O==1),
  BYA17P = ifelse(BYA17P<0, NA, BYA17P==1),
  BYA17Q = ifelse(BYA17Q<0, NA, BYA17Q==1),
  BYA17R = ifelse(BYA17R<0, NA, BYA17R==1),
  COURSE_OFFERING_SCORE = BYA17A +
  BYA17B +
  BYA17C+
  BYA17D +
  BYA17E+
  BYA17F+
  BYA17G+
  BYA17H +
  BYA17I +
  BYA17J +
  BYA17K +
  BYA17L +
  BYA17M +
  BYA17N +
  BYA17O+
  BYA17P +
  BYA17Q +
  BYA17R 
)  

hist(els$COURSE_OFFERING_SCORE)

els %>%
  filter(!is.na(COURSE_OFFERING_SCORE)) %>%
  ggplot() +
  geom_point(aes(y=BYTXCSTD, x=COURSE_OFFERING_SCORE))

els %>%
  filter(!is.na(COURSE_OFFERING_SCORE)) %>%
  ggplot() +
  geom_point(aes(y=F3ERN2011, x=COURSE_OFFERING_SCORE))

els %>%
  filter(!is.na(COURSE_OFFERING_SCORE), !is.na(ED_ATNMT)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT, y=COURSE_OFFERING_SCORE))
```


good teachers get higher pay
BYA28F


F1A17D         14466-14468   Tutoring/individualized academic program if fail competency test


F1A29DP        14599-14601   # of full-time teachers whose highest degree is EdD or PhD
F1A30          14606-14609   # of full-time library media resource center staff members
F1A35          14647-14650   % of full-time library staff is state-certified

BYL11AA        14792-14794   Library has telephone
BYL11CA        14812-14814   Library has photocopier

BYL03F         14752-14754   Library has computer access area/lab


```{r}
els <- els %>% mutate(
  TUTORING = ifelse(F1A17D < 0, NA, F1A17D),
  TUTORING = case_when(
    TUTORING == 1 ~ 0,
    TUTORING == 2 | TUTORING==3 ~ 1
  ),
  TUTORING= factor(TUTORING)
)



els <- els %>% mutate(
  ED_ATNMT_1 = case_when(
    F3ATTAINMENT == 1 ~ "No GED",
    F3ATTAINMENT == 2 ~ "GED",
    F3ATTAINMENT ==3 | F3ATTAINMENT ==4 | F3ATTAINMENT ==5 ~ "Associates/Some Undergrad",
    F3ATTAINMENT ==6 ~ "Bachelors",
    F3ATTAINMENT ==7 | F3ATTAINMENT ==8 | F3ATTAINMENT ==10~ "Futher Education"
  )
)
els$ED_ATNMT_1 = factor(els$ED_ATNMT_1, levels=c( "No GED", "GED", "Associates/Some Undergrad", "Bachelors", "Further Education"))

els %>%
  filter(!is.na(TUTORING)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=TUTORING)) +
  facet_grid(.~SESQU)

els %>%
  filter(!is.na(TUTORING)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=TUTORING)) +
  facet_grid(.~SESQU)

els %>%
  filter(!is.na(TUTORING), !is.na(ED_ATNMT), !is.na(SESQU)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT_1, fill=TUTORING), position = "fill") +
  facet_grid(.~SESQU)


```

```{r}
els <- els %>% mutate(
  GOOD_TEACHERS_HIGHER_PAY = ifelse(BYA28F < 0, NA, BYA28F),
  GOOD_TEACHERS_HIGHER_PAY= factor(GOOD_TEACHERS_HIGHER_PAY)
)

els %>%
  filter(!is.na(GOOD_TEACHERS_HIGHER_PAY)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=GOOD_TEACHERS_HIGHER_PAY)) +
  facet_grid(.~SESQU)

els %>%
  filter(!is.na(GOOD_TEACHERS_HIGHER_PAY)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=GOOD_TEACHERS_HIGHER_PAY)) +
  facet_grid(.~SESQU)

els %>%
  filter(!is.na(GOOD_TEACHERS_HIGHER_PAY), !is.na(ED_ATNMT), !is.na(SESQU)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT_1, fill=GOOD_TEACHERS_HIGHER_PAY), position = "fill") +
  facet_grid(.~SESQU)


```

BYA42N         14183-14185   Teachers have access to Internet professional development programs
BYA43A         14185-14187   Teacher training on use of new software
BYA43B         14187-14189   Teacher training on use of Internet
BYA43C         14189-14191   Teacher training on using computers to teach skills
BYA43D         14191-14193   Teacher training on integrating computer into class
BYA43E         14193-14195   Teacher training on basic computer literacy
```{r}
els <- els %>% mutate(
  TEACHERS_INTERNET_PROGRAM = ifelse(BYA42N < 0, NA, BYA42N),
  BYA43A = ifelse(BYA43A < 0, NA, BYA43A),
  BYA43B = ifelse(BYA43B < 0, NA, BYA43B),
  BYA43C = ifelse(BYA43C < 0, NA, BYA43C),
  BYA43D = ifelse(BYA43D < 0, NA, BYA43D),
  BYA43E = ifelse(BYA43E < 0, NA, BYA43E),

  TEACHER_TRAINING_TECH= factor(BYA43A+BYA43B+BYA43C+BYA43D+BYA43E)
)


els %>%
  filter(!is.na(TEACHER_TRAINING_TECH)) %>%
  ggplot() +
  geom_boxplot(aes(y=BYTXCSTD, x=TEACHER_TRAINING_TECH))

els %>%
  filter(!is.na(TEACHER_TRAINING_TECH)) %>%
  ggplot() +
  geom_boxplot(aes(y=F3ERN2011, x=TEACHER_TRAINING_TECH)) 

els %>%
  filter(!is.na(TEACHER_TRAINING_TECH), !is.na(ED_ATNMT), !is.na(SESQU)) %>%
  ggplot() +
  geom_bar(aes(x=ED_ATNMT_1, fill=TEACHER_TRAINING_TECH), position = "fill")


```


BYA14B         13749-13752   % 10th-graders in college prep program
BYA15C         13783-13785   Students in program to prepare for college

```{r}
els %>%
  ggplot() +
  geom_point(aes(x=BYA14B, y=F3ERN2011 ))

els %>%
  filter(!is.na(BYA14B), !is.na(ED_ATNMT), !is.na(SESQU)) %>%
  ggplot() +
  geom_boxplot(aes(x=ED_ATNMT_1, y=BYA14B)) +
  facet_grid(.~SESQU)

```

