---
title: "Results"
author: "Christina Chen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(MASS)
library(lme4)
library(mice)

# data file too large to push to git - split into two halfs in split_data.R
student_data1 = read.csv("els_student_first_half.csv")
student_data2 = read.csv("els_student_second_half.csv")
els = rbind(student_data1, student_data2)
els_original=els
```


```{r}

els <- els %>%
  mutate(TEST_COMP_QU = ifelse(BYTXCQU < 0, NA, BYTXCQU),
         TEST_COMP = ifelse(BYTXCSTD < 0, NA, BYTXCSTD)
         )


els <- els %>%
  mutate(SES_2011 = ifelse(F3SES < 0, NA, F3SES),
         SESQU_2011 = ifelse(F3SESQU < 0, NA, F3SESQU), SESQU_2011=factor(SESQU_2011, levels=c(1,2,3,4)))

els <- els %>%
  mutate(TEST_QU = ifelse(BYTXCQU < 0, NA, BYTXCQU), TEST_QU=factor(TEST_QU))

els <- els %>% mutate(
  ED_ATNMT = case_when(
    F3ATTAINMENT == 1 ~ "No GED",
    F3ATTAINMENT == 2 ~ "GED",
    F3ATTAINMENT ==3 | F3ATTAINMENT ==4 | F3ATTAINMENT ==5 ~ "Associates/\nSome Undergrad",
    F3ATTAINMENT ==6 ~ "Bachelors",
    F3ATTAINMENT ==7 | F3ATTAINMENT ==8~ "Master's/\nPost-Bacc. Cert.",
    F3ATTAINMENT ==10 ~ "PhD"
  )
)
els$ED_ATNMT = factor(els$ED_ATNMT, levels=c( "No GED", "GED", "Associates/\nSome Undergrad", "Bachelors", "Master's/\nPost-Bacc. Cert.", "PhD"))

```

```{r}
# AI/AN = American Indian/Alaska Native
# API = Asian, Hawaii/Pacific Islander
els <- els %>% mutate(
  RACE = case_when(
    BYRACE == 1 | BYRACE ==6~ "AI/AN/Mixed",
    BYRACE ==2 ~ "API",
    BYRACE ==3 ~ "Black",
    BYRACE ==4 | BYRACE ==5 ~ "Hispanic",
    BYRACE ==7 ~ "White"
  )
)
els$RACE = factor(els$RACE, levels=c( "AI/AN/Mixed", "API", "Black", "Hispanic","White"))


# used base year data, if na, used f1 data
els <- els %>% mutate(
  BYFCOMP = na_if(BYFCOMP, BYFCOMP<0),
  FAM_COMP = case_when(
    BYFCOMP == 1 | BYFCOMP == 2 | BYFCOMP == 3 | BYFCOMP == 4 ~ "Two Parents",
    BYFCOMP ==5 | BYFCOMP == 6 | BYFCOMP ==7 | BYFCOMP == 8 | BYFCOMP == 9~ "Single/Absent Parent",
    F1FCOMP == 1 | F1FCOMP == 2 | F1FCOMP == 3 | F1FCOMP == 4~ "Two Parents",
    F1FCOMP ==5 | F1FCOMP == 6 | F1FCOMP ==7 | F1FCOMP == 8 | F1FCOMP == 9~ "Single/Absent Parent"
  )
)

els <- els %>% mutate(
  BYPARED = na_if(BYPARED, BYPARED<0),
  PT_ED = case_when(
    BYPARED == 1 | BYPARED == 2 ~ "Some HS/GED",
    BYPARED == 3 | BYPARED == 4| BYPARED == 5 ~ "Attend PS/Associate's",
    BYPARED == 6 ~ "Bachelors",
    BYPARED == 7 | BYPARED == 8 ~ "Advanced Degree"
  ),
  PT_ED_YEARS = case_when(
    BYPARED == 1 ~ 1,
    BYPARED == 2 ~ 4,
    BYPARED == 3 ~ 5,
    BYPARED == 4 ~ 6,
    BYPARED == 5 ~ 7,
    BYPARED == 6 ~ 8,
    BYPARED == 7 ~ 10,
    BYPARED == 8 ~ 12
  )
)
els$PT_ED = factor(els$PT_ED, levels=c("Some HS/GED", "Attend PS/Associate's","Bachelors", "Advanced Degree"))
els$PT_ED_YEARS = as.numeric(els$PT_ED_YEARS)


els <- els %>% mutate(
  SESQU = ifelse(BYSES1QU < 0, NA, BYSES1QU),
  SESQU = case_when(
    SESQU == 1 ~ "1st SES Quartile",
    SESQU == 2 ~ "2nd SES Quartile",
    SESQU == 3 ~ "3rd SES Quartile",
    SESQU == 4 ~ "4th SES Quartile"
  ),
  SESQU=factor(SESQU),
  SES = ifelse(BYSES1 <= -2, NA, BYSES1)
)

# els <- els %>% mutate(FAM_INCOME = factor(BYINCOME))

els <- els %>% mutate(
  PT_PUSH_1 = na_if(BYPARASP, BYPARASP<0),
  PT_PUSH = case_when(
    PT_PUSH_1 == 1 | PT_PUSH_1 ==2 ~ "Some HS/GED",
    PT_PUSH_1 == 3 | PT_PUSH_1 == 4 ~ "Attend PS/Associate's",
    PT_PUSH_1 == 5 ~ "Bachelor's",
    PT_PUSH_1 == 6 | PT_PUSH_1 == 7~ "Advanced Degree"
  ),
  PT_PUSH_YEARS = case_when(
    PT_PUSH_1 == 1 ~ 2,
    PT_PUSH_1 == 2 ~ 4,
    PT_PUSH_1 == 3 ~ 6,
    PT_PUSH_1 == 4 ~ 7,
    PT_PUSH_1 == 5 ~ 8,
    PT_PUSH_1 == 6 ~ 10,
    PT_PUSH_1 == 7 ~ 12
  )
)
els$PT_PUSH = factor(els$PT_PUSH, levels=c("Some HS/GED", "Any college (2 or 4 year)", "Bachelor's","Advanced Degree"))
els$PT_PUSH_YEARS = as.numeric(els$PT_PUSH_YEARS)


els <- els %>% mutate(
  COMP_HOME = ifelse(BYS84C < 0, NA, BYS84C),
  COMP_HOME=factor(COMP_HOME)
)

els <- els %>% mutate(
  HAS_INTERNET = ifelse(BYS84D < 0, NA, BYS84D),
  HAS_INTERNET=factor(HAS_INTERNET),
  
  COMP_AND_INTERNET = HAS_INTERNET == 1 & COMP_HOME == 1
)

```



```{r}
els <- els %>% mutate(
  SCHOOL_TYPE=case_when(
    BYSCTRL == 1 ~ "Public",
    BYSCTRL == 2 ~ "Catholic",
    BYSCTRL == 3 ~ "Private"
  )
)

els <- els %>% mutate(
  SCHOOL_URBAN=case_when(
    BYURBAN == 1 ~ "Urban",
    BYURBAN == 2 ~ "Suburban",
    BYURBAN == 3 ~ "Rural"
  )
)

els <- els %>% mutate(
  DROPOUT_PREV_PGM = ifelse(F1A23 < 0, NA, F1A23),
  DROPOUT_PREV_PGM=factor(DROPOUT_PREV_PGM)
)

els <- els %>% mutate(
  PRG_PGM = ifelse(F1A21F < 0, NA, F1A21F),
  PRG_PGM=factor(PRG_PGM)
)

els <- els %>% mutate(
  VOCATIONAL_PGM = ifelse(F1A21A < 0, NA, F1A21A),
  VOCATIONAL_PGM=factor(VOCATIONAL_PGM)
)


els <- els %>% mutate(
  BYA50A_1 = ifelse(BYA50A<0, NA, BYA50A),
  BYA50B_1 = ifelse(BYA50B<0, NA, BYA50B),
  BYA50C_1 = ifelse(BYA50C<0, NA, BYA50C),
  BYA50D_1 = ifelse(BYA50D<0, NA, BYA50D),
  BYA50E_1 = ifelse(BYA50E<0, NA, BYA50E),
  BYA50F_1 = ifelse(BYA50F<0, NA, BYA50F),
  BYA50G_1 = ifelse(BYA50G<0, NA, BYA50G),
  BYA50H_1 = ifelse(BYA50H<0, NA, BYA50H),
  BYA50I_1 = ifelse( BYA50I<0,NA, BYA50I),
  BYA50J_1 = ifelse( BYA50J<0, NA, BYA50J),
  BYA50K_1 = ifelse(BYA50K<0, NA, BYA50K),
  LH_LACK_OF_SPACE = factor(case_when(
    BYA50E_1 <= 1 ~ "No",
    BYA50E_1 <= 4 ~ "Yes"
  ), levels=c("No",  "Yes")),
  LH_POOR_BUILDINGS = factor(case_when(
    BYA50A_1 <= 1 ~ "No",
    BYA50A_1 <= 4 ~ "Yes"
  ), levels=c("No",  "Yes")),
  LH_POOR_HEATING_AIR_LIGHT = factor(case_when(
    BYA50B_1 <= 1 ~ "No",
    # BYA50B_1 <= 2 ~ "Some",
    BYA50B_1 <= 4 ~ "Yes"
  ), levels=c("No", "Yes")),
  LH_LACK_TEXT = factor(case_when(
    BYA50G_1 <= 1 ~ "No",
    BYA50G_1 <= 4 ~ "Yes"
  ), levels=c("No",  "Yes")),
  LH_LACK_COMPUTERS = factor(case_when(
    BYA50H_1 <= 1 ~ "No",
    BYA50H_1 <= 4 ~ "Yes"
  ), levels=c("No", "Yes")),
  LH_LACK_TECH_FACILITIES = factor(case_when(
    BYA50K_1 <= 1 ~ "No",
    BYA50K_1 <= 4 ~ "Yes"
  ), levels=c("No", "Yes")),
  LH_POOR_LIB = factor(case_when(
    BYA50F_1 <= 1 ~ "No",
    BYA50F_1 <= 4 ~ "Yes"
  ), levels=c("No", "Yes")),
  LH_POOR_SCIENCE_LAB = factor(case_when(
    BYA50C_1 <= 1 ~ "No",
    BYA50C_1 <= 4 ~ "Yes"
  ), levels=c("No","Yes")),
  LH_POOR_ARTS_FACILITIES = factor(case_when(
    BYA50D_1 <= 1 ~ "No",
    BYA50D_1 <= 4 ~ "Yes"
  ), levels=c("No", "Yes")),
  LH_POOR_MULTIMEDIA = factor(case_when(
    BYA50I_1 <= 1 ~ "No",
    BYA50I_1 <= 4 ~ "Yes"
  ), levels=c("No", "Yes")),
  
  LH_POOR_FACILITIES = LH_POOR_ARTS_FACILITIES=="Yes" | LH_POOR_SCIENCE_LAB=="Yes" | LH_POOR_LIB == "Yes",
  LH_POOR_TECH = LH_LACK_TECH_FACILITIES == "Yes" | LH_POOR_MULTIMEDIA == "Yes" | LH_LACK_COMPUTERS == "Yes",
  LH_POOR_BUILDINGS_AIR = LH_POOR_BUILDINGS == "Yes" | LH_POOR_HEATING_AIR_LIGHT == "Yes"
  
  # LEARNING_HINDERED_POOR_CONDITIONS_SCORE = BYA50A_1 + BYA50B_1 + BYA50E_1,
  # LEARNING_HINDERED_LACK_SUPPLIES_SCORE = BYA50G_1,
  # LEARNING_HINDERED_POOR_TECH_SCORE = BYA50D_1 + BYA50I_1 + BYA50H_1,
  # LEARNING_HINDERED_LACK_SPACE_SCORE = BYA50E_1,
  # LEARNING_HINDERED_POOR_FACILITIES_SCORE = BYA50D_1+BYA50C_1+BYA50F_1
)

els <- els %>% mutate(
  FREE_LUNCH_PERC = factor(ifelse(F1SCFLP < 0, NA, F1SCFLP)),
  FREE_LUNCH_PERC_10 = factor(ifelse(BY10FLP < 0, NA, BY10FLP))

)


els <- els %>% mutate(
  SECURITY = factor(ifelse(BYA40A < 0, NA, BYA40A)),
  FREE_LUNCH_CATEGORIES = case_when(
    as.numeric(FREE_LUNCH_PERC) == 1 ~ "<=5%",
    as.numeric(FREE_LUNCH_PERC) <= 4 ~ "<=30%",
    as.numeric(FREE_LUNCH_PERC) <= 5 ~ "<=50%",
    as.numeric(FREE_LUNCH_PERC) <= 6 ~ "<=75%",
    as.numeric(FREE_LUNCH_PERC) <= 7 ~ ">75%",
    
    as.numeric(FREE_LUNCH_PERC_10) == 1 ~ "<=5%",
    as.numeric(FREE_LUNCH_PERC) <= 4 ~ "<=30%",
    as.numeric(FREE_LUNCH_PERC_10) <= 5 ~ "<=50%",
    as.numeric(FREE_LUNCH_PERC_10) <= 6 ~ "<=75%",
    as.numeric(FREE_LUNCH_PERC_10) <= 7 ~ ">75%"
    
  )

)

# els <- els %>% mutate(
#   FREE_LUNCH_CATEGORIES = case_when(
#     as.numeric(FREE_LUNCH_PERC) <= 1 ~ "<=5%",
#     as.numeric(FREE_LUNCH_PERC) <= 4 ~ "<=30%",
#     as.numeric(FREE_LUNCH_PERC) <= 5 ~ "<=50%",
#     # as.numeric(FREE_LUNCH_PERC) <= 6 ~ "<=75%",
#     as.numeric(FREE_LUNCH_PERC) <= 7 ~ ">50%"
#   )
# 
# )

# els$FREE_LUNCH_CATEGORIES = factor(els$FREE_LUNCH_CATEGORIES, levels = c("<=5%",  "<=30%", "<=50%", ">50%"))
els$FREE_LUNCH_CATEGORIES = factor(els$FREE_LUNCH_CATEGORIES, levels = c("<=5%", "<=30%", "<=50%", "<=75%", ">75%"))

# assume negative value is 0
els <- els %>% mutate(
  BYA41A_1 = ifelse( BYA41A<0, 0, BYA41A),
  BYA41B_1 = ifelse( BYA41B<0, 0, BYA41B),
  BYA41C_1 = ifelse( BYA41C<0, 0, BYA41C),
  BYA41D_1 = ifelse( BYA41D<0, 0, BYA41D),
  BYA41E_1 = ifelse( BYA41E<0, 0, BYA41E),
  BYA41F_1 = ifelse( BYA41F<0, 0, BYA41F),
  BYA41G_1 = ifelse( BYA41G<0, 0, BYA41G),
  BYA41H_1 = ifelse( BYA41H<0, 0, BYA41H),
  BYA41I_1 = ifelse( BYA41I<0, 0, BYA41I),
  BYA41J_1 = ifelse( BYA41J<0, 0, BYA41J),
  BYA41K_1 = ifelse( BYA41K<0, 0, BYA41K),
  BYA41L_1 = ifelse( BYA41L<0, 0, BYA41L),
  BYA41M_1 = ifelse(BYA41M<0, 0, BYA41M),
  TEACHER_BASIC_TECH_ACCESS = factor(BYA41K_1 & BYA41L_1 & BYA41M_1),
  TEACHER_TECH_ACCESS = BYA41A_1+ BYA41B_1+BYA41C_1+BYA41D_1+BYA41E_1+BYA41F_1+BYA41G_1+BYA41H_1+BYA41I_1+BYA41J_1+2*BYA41K_1+2*BYA41L_1+ BYA41M_1
)

els <- els %>% mutate(
  BYA43A = ifelse(BYA43A < 0, NA, BYA43A),
  BYA43B = ifelse(BYA43B < 0, NA, BYA43B),
  BYA43C = ifelse(BYA43C < 0, NA, BYA43C),
  BYA43D = ifelse(BYA43D < 0, NA, BYA43D),
  BYA43E = ifelse(BYA43E < 0, NA, BYA43E),

  TEACHER_TRAINING_TECH= BYA43A+BYA43B+BYA43C+BYA43D+BYA43E
)


els <- els %>% mutate(
  GOOD_TEACHERS_HIGHER_PAY = ifelse(BYA28F < 0, NA, BYA28F),
  GOOD_TEACHERS_HIGHER_PAY= factor(GOOD_TEACHERS_HIGHER_PAY)
)

els <- els %>% mutate(
  TUTORING = ifelse(F1A17D < 0, NA, F1A17D),
  TUTORING = case_when(
    TUTORING == 1 ~ 0,
    TUTORING == 2 | TUTORING==3 ~ 1
  ),
  TUTORING= factor(TUTORING)
)

els <- els %>% mutate(
  LIBRARY_COMPUTER_LAB = ifelse(BYL03F < 0, NA, BYL03F),
  LIBRARY_COMPUTER_LAB= factor(LIBRARY_COMPUTER_LAB)
)

els <- els %>% mutate(
  HIGHEST_TEACHER_SALARY = ifelse(BYA26B < 0, NA, BYA26B),
  LOWEST_TEACHER_SALARY = ifelse(BYA26A < 0, NA, BYA26A),
  TEACHER_SALARY_DISPARITY = HIGHEST_TEACHER_SALARY - LOWEST_TEACHER_SALARY,
  LOWEST_TEACHER_SALARY_THOUSANDS = LOWEST_TEACHER_SALARY/1000
)
els$PERC_SOPH_COLLEGE_PREP = els$BYA14B
```


```{r}
els <- els %>%
  mutate(
   DROPPED_OUT = case_when(
     F3EVERDO == 1 ~ TRUE,
     F3EVERDO == 0 ~ FALSE
   )
)
els <- els %>%
  mutate(ED_ATNMT_BACHELORS = !(ED_ATNMT == 'No GED' | ED_ATNMT == 'GED' | ED_ATNMT == 'Associates/\nSome Undergrad')
)

els <- els %>%
  mutate(
    SCHOOL_TYPE1 = SCHOOL_TYPE,
    SCHOOL_TYPE = case_when(
      SCHOOL_TYPE == "Private" | SCHOOL_TYPE =="Catholic" ~ "Private/Catholic",
      SCHOOL_TYPE == "Public" ~ "Public"
    )
  )

els <- els %>%
  mutate(FREE_LUNCH_MAJ = 
    !(FREE_LUNCH_CATEGORIES == "<=30%" |  FREE_LUNCH_CATEGORIES == "<=5%" | FREE_LUNCH_CATEGORIES == "<=50%")
  )

els_new <- els[,c("SCHOOL_TYPE", "RACE", "FAM_COMP", "SES", "PT_ED", "PT_PUSH", "COMP_AND_INTERNET", "TEACHER_TECH_ACCESS",  "PERC_SOPH_COLLEGE_PREP", "LOWEST_TEACHER_SALARY_THOUSANDS", "TEST_COMP", "SESQU_2011", "DROPPED_OUT", "ED_ATNMT_BACHELORS", "PT_ED_YEARS", "PT_PUSH_YEARS", "SES_2011", 
                  "LH_LACK_OF_SPACE", "LH_POOR_BUILDINGS", "LH_POOR_HEATING_AIR_LIGHT", "LH_LACK_TEXT", "LH_POOR_FACILITIES", "LH_POOR_TECH", "FREE_LUNCH_CATEGORIES", "FREE_LUNCH_MAJ"
  )]

# https://stats.idre.ucla.edu/r/faq/how-do-i-perform-multiple-imputation-using-predictive-mean-matching-in-r/
els_imputed <- mice(els_new, m=1, maxit = 50, seed = 500, method="pmm")

```

```{r}

els_complete <- complete(els_imputed, 1)
```

# Results

Due to the number of variables used in the models, there are many observations that have missing data in at least one of the variables. Therefore, we use the `mice` package to impute the data using the predictive mean-matching method.

## Standardized Test Composite Score

Our first measure of student success is standardized test composite score, which combines the math and reading scores. In order to compare the correlations between family and school predictors and the test composite, we are using a model that includes predictors from both groups. This way, the effect sizes of variables of interest are adjusted for by the inclusion of more potential confounding variables. Since test composite is a continuous variable, we decided to an ordinary least squares model specified below:


\begin{align*}
\text{test composite}_i = &\beta_0 + \beta_11(\text{school type=private}) + \beta_21(\text{school type=public}) +

&\beta_3 1(\text{race=Mixed}) + \beta_4 1(\text{race=API}) + \beta_5 1(\text{race=Black}) + 
\beta_{6} 1(\text{race=Hispanic}) + \beta_{7} 1(\text{race=White}) +
\beta_{8} 1(\text{family composition=two parents}) +  \beta_{9} \text{SES} \\+
&\beta_{10}(\text{parents' # years ed.}) + \beta_{13}(\text{# years ed. parents desire for student}) + \beta_{16}1(\text{has computer at home}) + \beta_{14}1(\text{has internet at home}) \\
+ &\beta_{15}(\text{teachers' tech access score}) + \beta_{16}1(\text{library has computer lab})
+\beta_{17}(\text{% sophomores in college prep program}) + \beta_{18}(\text{lowest teacher salary (thousands)}) \\
+ & \beta_{19}(\text{learning hindered by poor buildings}) + \beta_{20}(\text{learning hindered by poor heating/air/light}) + \beta_{21}1(\text{learning hindered by poor science labs}) \\+& \beta_{22}1(\text{learning hindered by poor fine arts facilities})+ \beta_{23}1(\text{learning hindered by lack of space})+ \beta_{24}1(\text{learning hindered by poor library})  \\
+& \beta_{25}1(\text{learning hindered by lack of text/supplies}) + \beta_{26}1(\text{learning hindered by too few computers}) \\
+& \beta_{27}1(\text{learning hindered by lack of tech equipment})+ \beta_{28}1(\text{free lunch percentage} > 50%) + \epsilon_i, \ 
\text{where } \epsilon_i \stackrel{i.i.d.}{\sim} \mathcal{N}(0, \sigma^2)
\end{align*}


OLS models have model assumptions of independence, normality, homoscedasticity (constant variance), and linearity between predictors and the response. Since each of the observations in our data comes from a single student, the independence assumption is satisfied. To assess the other assumptions and multicollinearity, model diagnostics are analyzed in the Appendix.

```{r}

test_model_both = lm(TEST_COMP ~ SCHOOL_TYPE  + RACE + FAM_COMP + SES + PT_ED_YEARS  + PT_PUSH_YEARS  + COMP_AND_INTERNET + TEACHER_TECH_ACCESS + PERC_SOPH_COLLEGE_PREP + LOWEST_TEACHER_SALARY_THOUSANDS +
                       LH_LACK_OF_SPACE*FREE_LUNCH_MAJ + LH_POOR_BUILDINGS+ LH_POOR_HEATING_AIR_LIGHT + LH_LACK_TEXT + LH_POOR_FACILITIES + LH_POOR_TECH, data = els_complete)
summary(test_model_both)
# View(els[15110,c("SCHOOL_TYPE", "SCHOOL_URBAN","DROPOUT_PREV_PGM","VOCATIONAL_PGM","POOR_BUILDINGS","POOR_HEATING_AIR_LIGHT","LACK_TEXT","FREE_LUNCH_CATEGORIES","LACK_OF_SPACE","SECURITY", "RACE","FAM_COMP","SES","PT_ED", "PT_PUSH","COMP_HOME","HAS_INTERNET")])


# SENSITIVITY ANALYSSI
# test_model_both = lm(TEST_COMP ~ SCHOOL_TYPE  + RACE + FAM_COMP + SES + PT_ED_YEARS  + PT_PUSH_YEARS  + COMP_HOME + HAS_INTERNET + TEACHER_TECH_ACCESS + PERC_SOPH_COLLEGE_PREP + LOWEST_TEACHER_SALARY_THOUSANDS +BYA50A_1*FREE_LUNCH_MAJ+
#   BYA50B_1+
#   BYA50C_1+
#   BYA50D_1+
#   BYA50E_1+
#   BYA50F_1+
#   BYA50G_1+
#   BYA50H_1+
#   BYA50K_1, data = els_complete)
# summary(test_model_both)
```


## Socioeconomic Status
```{r}
# https://stats.idre.ucla.edu/r/dae/ordinal-logistic-regression/
# ses_model_family = polr(SESQU_2011 ~ RACE + FAM_COMP + SES + PT_ED + PT_PUSH + COMP_HOME + HAS_INTERNET, data = els, Hess = TRUE)
# ses_model_family_summary = summary(ses_model_family)
# (ctable <- coef(ses_model_family_summary))
# p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
# (ctable <- cbind(ctable, "p value" = p))

# https://stats.idre.ucla.edu/r/dae/ordinal-logistic-regression/
ses_model = polr(SESQU_2011 ~ SCHOOL_TYPE  + RACE + FAM_COMP + SES + PT_ED_YEARS  + PT_PUSH_YEARS  + COMP_AND_INTERNET + TEACHER_TECH_ACCESS + PERC_SOPH_COLLEGE_PREP + LOWEST_TEACHER_SALARY_THOUSANDS +
                       LH_LACK_OF_SPACE*FREE_LUNCH_MAJ + LH_POOR_BUILDINGS+ LH_POOR_HEATING_AIR_LIGHT + LH_LACK_TEXT + LH_POOR_FACILITIES + LH_POOR_TECH, data = els_complete, Hess = TRUE)
ses_model_summary = summary(ses_model)
ctable <- coef(ses_model_summary)
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
is_sig <- p <= 0.05
(ctable <- cbind(ctable, "p value" = p, "sig"=is_sig))


# test_model_both = lm(TEST_COMP ~ SCHOOL_TYPE+SCHOOL_URBAN+DROPOUT_PREV_PGM+VOCATIONAL_PGM  + RACE + FAM_COMP + SES + PT_ED + PT_PUSH + COMP_HOME + HAS_INTERNET + TEACHER_TECH_ACCESS + LIBRARY_COMPUTER_LAB + PERC_SOPH_COLLEGE_PREP + LOWEST_TEACHER_SALARY_THOUSANDS +LH_POOR_BUILDINGS+LH_POOR_HEATING_AIR_LIGHT+LH_LACK_TEXT+LH_LACK_OF_SPACE+LH_LACK_COMPUTERS+LH_LACK_TECH_FACILITIES+LH_POOR_LIB+LH_POOR_SCIENCE_LAB+LH_POOR_ARTS_FACILITIES + FREE_LUNCH_CATEGORIES, data = els_complete)
# summary(test_model_both)
# plot(test_model_both)

# sensitivy analysis with including test composite

```



```{r}
# ses_model_lm = lm(log(SES_2011+0.1) ~ SCHOOL_TYPE+SCHOOL_URBAN+DROPOUT_PREV_PGM+VOCATIONAL_PGM  + RACE + FAM_COMP + SES + PT_ED_YEARS + PT_PUSH_YEARS + COMP_HOME + HAS_INTERNET + TEACHER_TECH_ACCESS + LIBRARY_COMPUTER_LAB + PERC_SOPH_COLLEGE_PREP*FREE_LUNCH_CATEGORIES + LOWEST_TEACHER_SALARY_THOUSANDS +LH_POOR_BUILDINGS+LH_POOR_HEATING_AIR_LIGHT+LH_LACK_TEXT+LH_LACK_OF_SPACE+LH_LACK_COMPUTERS+LH_LACK_TECH_FACILITIES+LH_POOR_LIB+LH_POOR_SCIENCE_LAB+LH_POOR_ARTS_FACILITIES , data = els)
# 
# summary(ses_model_lm)
# plot(ses_model_lm)

```


## Education Attainment

Our third measure of success is the level of education the student attained. Since there are 6 levels that we coded from the data and our variables include lots of categorical variables, there is a high possibility of a small sample size for many of the groupings of levels between the predictors and outcome. Therefore, we analyze models for outcomes of whether a student graduated high school and whether a student graduated college using logistic regression models, defined below:

$$
\begin{align*}
logit( P(\text{Dropped Out HS}_i = Yes) = &\beta_0 + \beta_11(\text{school type=private}) + \beta_21(\text{school type=public}) +
+
&\beta_3 1(\text{race=Mixed}) + \beta_4 1(\text{race=API}) + \beta_5 1(\text{race=Black}) + 
\beta_{6} 1(\text{race=Hispanic}) + \beta_{7} 1(\text{race=White}) +
\beta_{8} 1(\text{family composition=two parents}) +  \beta_{9} \text{SES} \\+
&\beta_{10}(\text{parents' # years ed.}) + \beta_{13}(\text{# years ed. parents desire for student}) + \beta_{16}1(\text{has computer at home}) + \beta_{14}1(\text{has internet at home}) \\
+ &\beta_{15}(\text{teachers' tech access score})
+\beta_{17}(\text{% sophomores in college prep program}) + \beta_{18}(\text{lowest teacher salary (thousands)}) \\
+ & \beta_{19}(\text{learning hindered by poor buildings}) + \beta_{20}(\text{learning hindered by poor heating/air/light}) + \beta_{21}1(\text{learning hindered by poor science labs}) \\+& \beta_{22}1(\text{learning hindered by poor fine arts facilities})+ \beta_{23}1(\text{learning hindered by lack of space})+ \beta_{24}1(\text{learning hindered by poor library})  \\
+& \beta_{25}1(\text{learning hindered by lack of text/supplies}) + \beta_{26}1(\text{learning hindered by too few computers}) \\
+& \beta_{27}1(\text{learning hindered by lack of tech equipment})+\\

+& \beta_{31}1(\text{free lunch percentage}> 50%)\\

\end{align*}

\begin{align*}

logit( P(\text{Received Bachelor's}_i = Yes) = &\beta_0 + \beta_11(\text{school type=private}) + \beta_21(\text{school type=public}) +
+
&\beta_3 1(\text{race=Mixed}) + \beta_4 1(\text{race=API}) + \beta_5 1(\text{race=Black}) + 
\beta_{6} 1(\text{race=Hispanic}) + \beta_{7} 1(\text{race=White}) +
\beta_{8} 1(\text{family composition=two parents}) +  \beta_{9} \text{SES} \\+
&\beta_{10}(\text{parents' # years ed.}) + \beta_{13}(\text{# years ed. parents desire for student}) + \beta_{16}1(\text{has computer at home}) + \beta_{14}1(\text{has internet at home}) \\
+ &\beta_{15}(\text{teachers' tech access score}) + \beta_{16}1(\text{library has computer lab})
+\beta_{17}(\text{% sophomores in college prep program}) + \beta_{18}(\text{lowest teacher salary (thousands)}) \\
+ & \beta_{19}(\text{learning hindered by poor buildings}) + \beta_{20}(\text{learning hindered by poor heating/air/light}) + \beta_{21}1(\text{learning hindered by poor science labs}) \\+& \beta_{22}1(\text{learning hindered by poor fine arts facilities})+ \beta_{23}1(\text{learning hindered by lack of space})+ \beta_{24}1(\text{learning hindered by poor library})  \\
+& \beta_{25}1(\text{learning hindered by lack of text/supplies}) + \beta_{26}1(\text{learning hindered by too few computers}) \\
+& \beta_{27}1(\text{learning hindered by lack of tech equipment})+\\

+& \beta_{31}1(\text{free lunch percentage}> 50%)\\
\end{align*}

$$

The assumptions for logistic regression are independence, no multicollinearity, and a large sample size. All observations are independent, as they are sampled from individual students. Multicollinearity is tested for using VIF (see Appendix). Finally, since there are many categorical variables used in our model, there are many combinations of levels. Although categorical variable levels were combined to reduce the total number of samples needed, there are still a few levels that have only a couple of samples. Therefore, this assumption is not fully met due to imbalanced data. However, since we are only performing inference, this may not need to be addressed further, as we will not be analyzing the less frequent combinations of variables.

```{r}
# els_complete <- els_complete %>%
#   mutate(ED_ATNMT_CONDENSED = case_when(
#     ED_ATNMT == 'No GED' ~ 1,
#     ED_ATNMT == 'GED' | ED_ATNMT == 'Associates/\nSome Undergrad' ~ 2,
#     TRUE ~ 3
#   ))
# 
# els_complete$ED_ATNMT_CONDENSED = factor(els_complete$ED_ATNMT_CONDENSED, levels=c(1,2,3))



ed_model <- glm(DROPPED_OUT ~SCHOOL_TYPE  + RACE + FAM_COMP + SES + PT_ED_YEARS  + PT_PUSH_YEARS  + COMP_AND_INTERNET + TEACHER_TECH_ACCESS + PERC_SOPH_COLLEGE_PREP + LOWEST_TEACHER_SALARY_THOUSANDS +
                       LH_LACK_OF_SPACE*FREE_LUNCH_MAJ + LH_POOR_BUILDINGS+ LH_POOR_HEATING_AIR_LIGHT + LH_LACK_TEXT + LH_POOR_FACILITIES + LH_POOR_TECH, data = els_complete, family="binomial")
summary(ed_model)

```






<!-- ### Bachelor's Attained -->
```{r}

ed_model_bachelors <- glm(ED_ATNMT_BACHELORS ~ SCHOOL_TYPE  + RACE + FAM_COMP + SES + PT_ED_YEARS  + PT_PUSH_YEARS  + COMP_AND_INTERNET + TEACHER_TECH_ACCESS + PERC_SOPH_COLLEGE_PREP + LOWEST_TEACHER_SALARY_THOUSANDS +
                       LH_LACK_OF_SPACE*FREE_LUNCH_MAJ + LH_POOR_BUILDINGS+ LH_POOR_HEATING_AIR_LIGHT + LH_LACK_TEXT + LH_POOR_FACILITIES + LH_POOR_TECH, data = els_complete, family = "binomial")

summary(ed_model_bachelors)
```


# Appendix


## Standardized Test Composite Score

```{r}
plot(test_model_both)
car::vif(test_model_both)
```


## SES Status
```{r}
# proportional odds test
mlm <- nnet::multinom(SESQU_2011 ~ SCHOOL_TYPE+SCHOOL_URBAN+DROPOUT_PREV_PGM+VOCATIONAL_PGM  + RACE + FAM_COMP + SES + PT_ED_YEARS + PT_PUSH_YEARS + COMP_HOME + HAS_INTERNET + TEACHER_TECH_ACCESS + LIBRARY_COMPUTER_LAB + PERC_SOPH_COLLEGE_PREP*FREE_LUNCH_CATEGORIES + LOWEST_TEACHER_SALARY_THOUSANDS +LH_POOR_BUILDINGS+LH_POOR_HEATING_AIR_LIGHT+LH_LACK_TEXT+LH_LACK_OF_SPACE+LH_LACK_COMPUTERS+LH_LACK_TECH_FACILITIES+LH_POOR_LIB  + TEST_COMP, data = els_complete)
G <- -2 * (logLik(ses_model)[1] - logLik(mlm)[1])
pchisq(G, df = length(ses_model$zeta) - 1, lower.tail = FALSE)

car::vif(ses_model)
```


## Education Attainment
